Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-05 AO_guides/manuscriptAO.tex
Added youyu's comment about how surge enters domain from open water. Couldn't find the Greatbatch reference though

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly. Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the sea surface height anomaly. Fourth, a simulation with only wind forcing, that is no anomaly and no inverse barometer effect due to pressure. Comparisons between these simulations indicate the relative importance the the open boundary forcing, the local wind forcing, and the local atmospheric pressure forcing. Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction. Contributions due to stratification and bathymetric resolution will be left to a future study.

First, it is clear that the sea surface height anomaly at the open boundary contributes most significantly to the surge at each of these locations (Figure \ref{fig:factors}). When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations.  When atmospheric forcing is neglected, almost all of the surge amplitude is accounted for in the sea surface height anomaly forcing at the open boundary.

@@

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverese barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the importance of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation \ref{eq:inverse} or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more inline with the findings of \citep{murty1995storm} but does not alter the conclusion that the open boundary forcing contributes most significantly. Further, the model calculates the local atmopsheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

First, a comparison of the wind speed and direction for the two models and observations is provided in Figure \ref{fig:dec2012_weather} at Point Atkinson and at Sandheads, a lighthouse station in the southern Strait of Georgia. For each of the models, the wind speed and direction at the closest grid point to the longitude/latitude of Point Atkinson and Sandheads is used for comparison. Although the low resolution CGRF model has a grid spacing of 33 km, two grid points align closely with these locations. At Point Atkinson, the high resolution model is in very good agreement with observations for both the wind speed and direction, however, the low resolution CGRF model significantly underestimates the wind speed during high wind periods on Dec 15, 2012 and Dec 17, 2012. At Sandheads, both the low resolution and high resolution models have good agreement with observations. The low resolution model has a very good match at Sandheads likely because this grid cell happens to be mostly over water. Although not shown here, both models have a similar sea level atmospheric pressure at these two locations.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_weather.pdf}
\caption{Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson (upper) and Sandheads (lower). Observed winds shown in solid black, low resolution CGRF model in dashed black and high resolution model in grey. Observations are from the Environment Canada Climate Data archives \citep{ECClimateArchive}.}
\label{fig:dec2012_weather}
\end{figure}
%don't include direction?

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Hindcasts of five significant events in February 2006, November 2006, December 2006, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases and generally lower skill scores for the residuals.  Further, in a single simulation, the skill scores are fairly consistent between the four stations examined. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management.

@@

Finally, difficulties in reproducing accurate M$_2$ tidal harmonics in both the Strait of Georgia and the Strait of Juan de Fuca have been highlighted in this study among others \citep{stronach1993update, foreman2004m}. Future work will attempt to improve model accuracy by examining mixing parametrizations and energy dissipation due to bottom friction and their effect on tidal harmonics. Additionally, a biological-chemical modelling component will be added.

>>>>

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, atmospheric forcing, rivers, and sea surface height anomaly at the open boundary (or remote forcing). Second, a simulation with all forcing except the atmospheric forcing. Third, a simulation with all forcing except the remote forcing. Fourth, a simulation with only wind forcing, that is no remote forcing and no local atmospheric pressure. Comparisons between these simulations indicate the relative importance of the remote forcing, the local wind forcing, and the local atmospheric pressure forcing. Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction. Contributions due to stratification and bathymetric resolution will be left to a future study.

First, it is clear that the sea surface height anomaly at the open boundary contributes most significantly to the surge at each of these locations (Figure \ref{fig:factors}). When the sea surface height anomaly is not included as a forcing condition, the surge drops from a maximum 82 cm to 11 cm at Point Atkinson and its maximum occurs eight hours later.  A similar drop in the maximum surge and delay in timing of the maximum surge is observed at the other locations.  When atmospheric forcing is neglected, almost all of the surge amplitude is accounted for in remote forcing.

@@

A simple steady state argument balancing the wind stress with the pressure gradient can be used to calculate the slope of the sea surface in a body of water with a constant depth $D$ \citep{pugh2004changing}: 
\begin{equation}
\text{slope} = \frac{C\rho_{air}U^2}{g\rho D},
\end{equation}
where $C$ is the coefficient of friction, $\rho_{air}$ is the density of air, $g$ is the acceleration due to gravity, $U$ is the wind speed, and $\rho$ is the density of water. Given  $C=10^{-3}$, $\rho_{air}=1.23$ kg m$^{-3}$, $\rho= 1035$  kg m$^{-3}$, wind speed $U=20$ m s$^{-1}$, and depth $D=100$ m, the slope of the sea surface is approximately $5\times10^{-7}$. At Point Atkinson, a westerly wind can act over a small distance of approximately 50 km, giving a sea surface elevation on the order of centimetres in agreement with the model findings. Since the Strait of Georgia is not a constant depth and is up to 400 m deep, this calculation is only an approximation. Choosing a larger depth would result in an even smaller estimate for the sea surface elevation caused by wind stresses. Further, northwesterly winds would induce the largest elevations at Point Atkinson due to the large distance over which the winds can act.

%local pressure
Next, the surge amplitude decreases by a few centimetres at all four locations if local atmospheric pressure is not included in the surface forcing. Other findings have reported the inverse barometer effect due to atmospheric pressure contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}. Note that in these simulations the removal of the atmospheric pressure in the surface forcing does not entirely remove the inverese barometer effect from the model results since the sea surface height anomaly at the open boundary includes an inverse barometer component. As such, these simulations indicate the importance of the local atmospheric pressure forcing but not the inverse barometer entirely. The contribution due to the inverse barometer can be calculated with equation \ref{eq:inverse} or by the approximation that 1 hPa drop in pressure corresponds to 1 cm rise in sea level. In this event, the pressure at the Vancouver International Airport dropped to about 99 000 Pa \citep{ECClimateArchive}, corresponding to a 20 cm rise in sea level. This inverse barometer contribution is more inline with the findings of \citep{murty1995storm} but does not alter the conclusion that the remote forcing contributes most significantly. Further, the model calculates the local atmopsheric pressure forcing using a spatial gradient, and since the spatial scale of the low pressure system is large compared to the domain the local atmospheric pressure should not contribute greatly.

@@

First, a comparison of the wind speed and direction for the two models and observations is provided in Figure \ref{fig:dec2012_weather} at Point Atkinson and at Sandheads, a lighthouse station in the southern Strait of Georgia. For each of the models, the wind speed and direction at the closest grid point to the longitude/latitude of Point Atkinson and Sandheads is used for comparison. Although the low resolution CGRF model has a grid spacing of 33 km, two grid points align closely with these locations. At Point Atkinson, the high resolution model is in very good agreement with observations for both the wind speed and direction, however, the low resolution CGRF model significantly underestimates the wind speed during high wind periods on Dec 15, 2012 and Dec 17, 2012. At Sandheads, both the low resolution and high resolution models have good agreement with observations. The low resolution model has a very good match at Sandheads likely because this grid cell happens to be mostly over water. Although not shown here, both models have a similar sea level atmospheric pressure at these two locations. Note that the HRDPS model does not employ data assimilation.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_weather.pdf}
\caption{Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson (upper) and Sandheads (lower). Observed winds shown in solid black, low resolution CGRF model in dashed black and high resolution model in grey. Observations are from the Environment Canada Climate Data archives \citep{ECClimateArchive}.}
\label{fig:dec2012_weather}
\end{figure}

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmopsheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Hindcasts of five significant events in February 2006, November 2006, December 2006, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases and generally lower skill scores for the residuals.  Further, in a single simulation, the skill scores are fairly consistent between the four stations examined. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean, or remote forcing,  contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Effects due to stratification will also be considered in future work. Further, the remote forcing is applied uniformly across the open boundary and so any spatial differences in the incoming surge are not accounted for in the model.

@@

Finally, difficulties in reproducing accurate M$_2$ tidal harmonics in both the Strait of Georgia and the Strait of Juan de Fuca have been highlighted in this study among others \citep{stronach1993update, foreman2004m}. Future work will attempt to improve model accuracy by examining mixing parametrizations and energy dissipation due to bottom friction and their effect on tidal harmonics. Additionally, a biological-chemical modelling component will be added. Stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.
